---
output:
  ioslides_presentation:
    logo: logo.png
    smaller: yes
    widescreen: no
  beamer_presentation: default
  
---

## Packages we use


```{r echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr) 
library(magrittr)
library(tidyselect)
library(tidyverse)
library(datasets)
library(ggplot2)
```


## GSS census data

We read in the .csv-file. 

- the function `read_sav()` from package `haven` reads SPSS files.

- the function `read_stata` from the same package reads STATA .DTA files.

```{r echo=T, warning=FALSE, message=FALSE}

path <- "C:/Users/DstMove/Desktop/RGhana/GSS_data/CSV/"

cens10pct <- read.csv2(file = paste0("C:/Users/DstMove/Desktop/RGhana/GSS_data/CSV/",
                                     "defactopopn_10%_20221011d.csv"), sep = ",")

``` 

##

In order to decrease computation time, we select a 10 % sample of the 10 % sample for these exercises:

```{r echo=T, warning=FALSE, message=FALSE}

samprows <- sample(nrow(cens10pct), size = 0.10*nrow(cens10pct), replace = F)

samp10pct <- cens10pct[samprows,]

head(samp10pct)

``` 

## Reasons why people are not seeking work

First, do the calculation (we only count those who actually answered the question):

```{r echo=T, warning=FALSE, message=FALSE}

#Remove those without an answer
Wpop <- samp10pct[samp10pct$p13e != "",]

#Count the population
npop <- nrow(Wpop)

#Calculate percentages
No_work <- group_by(Wpop, p13e) %>% 
                       summarise(reasons = n(),
                       pct = reasons/npop*100)

``` 

##

Then, do the plot:

```{r echo=T, warning=FALSE, message=FALSE}

ggplot(data=No_work, aes(x=pct, y=p13e, fill=p13e)) +
  geom_bar(stat="identity") + ggtitle("Reasons for not seeking work") +
                              xlab("Percent")


``` 


## A plot of education by region

First, do the calculation:

```{r echo=T, warning=FALSE, message=FALSE}

edu_by_reg <- group_by(samp10pct, region) %>% 
                       summarise(all = n(),
                       educount = sum(p12a == "Attended earlier" | p12a == "Attending now", na.rm = T),
                       edupct = educount/all*100)

``` 

##

Then, do the plot (here, we have reversed the order of the legend using `guides` so that it's in the same order as the graph):

```{r echo=T, warning=FALSE, message=FALSE}

ggplot(data=edu_by_reg, aes(x=edupct, y=region,fill=region)) +
  geom_bar(stat="identity") + guides(fill = guide_legend(reverse = TRUE)) +
                              ggtitle("School Attendance by Region") +
                              xlab("Percentage ever attended school") 
                              


``` 


##

we could also do this as deviation from the norm:

```{r echo=T, warning=FALSE, message=FALSE}

edu_by_reg$meanpct <- mean(edu_by_reg$edupct)
edu_by_reg$edudev <- edu_by_reg$edupct - edu_by_reg$meanpct  
```

## Deviation from the norm-plot

```{r echo=T, warning=FALSE, message=FALSE}

ggplot(edu_by_reg, aes(x=region, y=edudev, fill = region)) + 
  guides(fill = guide_legend(reverse = TRUE)) +
  geom_bar(stat='identity', width=.75) + coord_flip()

```


## ICT device-ownership by type of tenancy

```{r echo=T, warning=FALSE, message=FALSE}

#Remove those without an answer
Hpop <- samp10pct[samp10pct$h05 != "",]

ICT_pop <- Hpop[!is.na(Hpop$p19atotal), ]

#Calculate percentages
ICT_own <- group_by(ICT_pop, p19atotal, h05) %>% 
           summarise(count = n()) 

ICT_own1 <- ICT_own

ICT_own1$percentages <- 0

ICT_own1[ICT_own1$p19atotal == 0,]$percentages <- ICT_own1[ICT_own1$p19atotal == 0,]$count/78460*100
ICT_own1[ICT_own1$p19atotal == 1,]$percentages <- ICT_own1[ICT_own1$p19atotal == 1,]$count/114908*100
ICT_own1[ICT_own1$p19atotal == 2,]$percentages <- ICT_own1[ICT_own1$p19atotal == 2,]$count/63530*100
ICT_own1[ICT_own1$p19atotal == 3,]$percentages <- ICT_own1[ICT_own1$p19atotal == 3,]$count/4622*100
ICT_own1[ICT_own1$p19atotal == 4,]$percentages <- ICT_own1[ICT_own1$p19atotal == 4,]$count/606*100

``` 

##

Then plot:

```{r echo=T, warning=FALSE, message=FALSE}

ggplot(ICT_own1, aes(fill=h05, y=percentages, x=p19atotal)) + 
    geom_bar(position="dodge", stat="identity")
```

##

We could also make it stacked:

```{r echo=T, warning=FALSE, message=FALSE}

ggplot(ICT_own, aes(fill=h05, y=count, x=p19atotal)) + 
    geom_bar(position="fill", stat="identity")
```

## Education by source of lightning:

A density plot

```{r echo=T, warning=FALSE, message=FALSE}
edu_light <- samp10pct %>% filter(restype == "Occupied housing unit") %>% group_by(h08a) %>% summarise(all = n(),
                              educount = sum(p12a == "Attended in the past" | p12a == "Attending now", na.rm = T),
                              edupct = educount/all*100)

ggplot(edu_light, aes(x=edupct, y=h08a)) + 
  guides(fill = guide_legend(reverse = T)) +
  geom_point(aes(col=h08a, size=edupct))

```

