---
title: 'BOP'
author: "Statistical Programming with R"
date: ""
output:
  ioslides_presentation:
    fig_height: 5
    fig_width: 8
    logo: logo_en.png
    transition: faster
    smaller: true
    df_print: tibble
  beamer_presentation: default
---
```{r setup, echo=F, results='hide', message = F, warning=F}
```


## Recreating the Maintenance and Repair BOP

- We received the BOP for Maintenance and Repair as an Excel-file
- First we inspect the dataset
  - Credit information in the sheet "Credit"
  - Debit information in the sheet "Debit"
  - Weights in the sheet "Ponder_Exp_Imp"
  - Output in sheet "PROVJERA"
- The credit and debit is already weighted, so we do not need the "Ponder_Exp_Imp"-sheet  

## Recreating the Maintenance and Repair BOP

Plan:

 - Read the sheets "Debit" and "Credit"
 - Calculate the balance for each country
 - Group the countries by area and summarize the credit, debit and balance
 - Write to excel

## Read the sheet - Credit

- The name of the sheet is "Credit"
- The data start from row 3 to 284
```{r warning = F}
library(openxlsx)
mar_credit <- read.xlsx("4_SB_MAINTENANCE and REPAIR.xlsx",
                        sheet = "Credit",
                        rows = 3:284) 
mar_credit
```


## Read the sheet - Debit

- The name of the sheet is "Debit"
- The data start from row 3 to 284
```{r}
mar_debit <- read.xlsx("4_SB_MAINTENANCE and REPAIR.xlsx",
                        sheet = "Debit",
                        rows = 3:284)
mar_debit
```

# Calculating the balance
## Tidy data
- The data is not tidy right now... Why?
- This makes it difficult to calculate the balance with tidyverse.
- And we need to load the tidyverse:
```{r message=F, warning = F}
library(tidyverse)
mar_credit 
```

## Tidy data
- We do not need description
- We want the dataset to be longer, with one row for each year
```{r}
mar_debit_long <- mar_debit |> 
  select(-Description) |> 
  pivot_longer(cols = matches("\\d{4}"),
               names_to = "year",
               values_to = "debit")
mar_debit_long
```

## Tidy data
- And the same for credit
```{r}
mar_credit_long <- mar_credit |> 
  select(-Description) |> 
  pivot_longer(cols = matches("\\d{4}"),
               names_to = "year",
               values_to = "credit")
mar_credit_long
```

## Combining the datasets
- We have two datasets that we need to combine to calculate the balance
- This is what a good place to use `joins`
```{r}
mar_long <- mar_credit_long |> 
  full_join(mar_debit_long,
            by = c("Code", "year"))
mar_long
```

## Calculating the balance
- Now it is straightforward to calculate the balance:
```{r}
mar_long <- mar_long |> 
  mutate(balance = credit-debit)
mar_long
```

## And datatypes...
- We notice that year is a character. We would prefer a numeric.
```{r}
mar_long <- mar_long |> 
  mutate(year = as.numeric(year))
mar_long
```

# Grouping data

## Groups
- Groupings into for instance
  - B6: EU27 (fixed composition) as of 31 January 2020 (brexit)
  - E1: Europe
  - E7: Central and Eastern Europe
- They are listed in "IZ Annex 1_BOP_DSD_Matrix_2-1_2022-09 - formatted_for_BOP_Vademecum" in the sheet "CL_AREA"
- Not in a very tidy format, so instead we use the file "groups.xlsx"

## Read the file
```{r}
eu_groups <- read.xlsx("groups.xlsx")
eu_groups
```

## Add it to our dataset

- Add the groups to our dataset from before
- The columns we join by is not spelled the same way
```{r warning = F, message = F}
group_long <- mar_long |> 
  left_join(eu_groups,
            by = c("Code" = "code"))
group_long
```
## Remove missing groups
- Some countries do not belong to a group!
```{r}
group_long |> 
  count(group)
```

## Remove missing groups
- Some countries do not belong to a group!
```{r}
group_long <- group_long |> 
  filter(!is.na(group))
group_long |> 
  count(group)
```

## Group by and summarise
- We want one row for every group for every year
- We want to sum the balance, debit and credit
```{r message = F}
group_long <- group_long |> 
  group_by(group, year) |> 
  summarise(debit = sum(debit),
            credit = sum(credit),
            balance = sum(balance)) |> 
  ungroup()
group_long
```
## Compare with target
- Compare the dataset with the expected result...
- Hopefully it matches the results you expected
- Small difference: Not called `group` but `Code`
```{r}
group_long <- group_long |> 
  rename(Code = group)
group_long
```


# Export to excel
## Untidy your data
- Look at the Excel-sheet we want to reproduce
- Two things differ:
  - One row for credit, debit and balance
  - One column for each year
  
## Untidy your data
- One row for credit, debit and balance.
- A longer dataset
```{r}
mar_untidy <- mar_long |> 
  pivot_longer(cols = c(credit, debit, balance),
               names_to = "bop",
               values_to = "value") 
mar_untidy
```
## Untidy your data
- Change values to C, D and B
```{r}
mar_untidy <- mar_untidy |> 
  mutate(bop = case_when(bop == "credit" ~ "C",
                         bop == "debit" ~ "D",
                         bop == "balance" ~ "B"))
mar_untidy
```
## Untidy your data
- One column for each year
- This makes our dataset wider
```{r}
mar_untidy <- mar_untidy |> 
  pivot_wider(names_from = year, values_from = value)
mar_untidy
```

## Untidy your data
- Do the same for the grouped data! Try doing it as one long pipe
- It should look like this:
```{r echo=F}
group_untidy <- group_long |> 
  pivot_longer(cols = c(credit, debit, balance),
               names_to = "bop",
               values_to = "value") |> 
  mutate(bop = case_when(bop == "credit" ~ "C",
                         bop == "debit" ~ "D",
                         bop == "balance" ~ "B")) |> 
  pivot_wider(names_from = year, values_from = value)
group_untidy
```


# Export to Excel
## Write to Excel-file
- We are already experts in this!
```{r}
library(openxlsx)
write.xlsx(mar_untidy, file = "mar_countries.xlsx")
write.xlsx(group_untidy, file = "mar_groups.xlsx")
```

## Write to Excel-file...
- So let's explore if we can do it a little bit better:
  - Both datasets in same file but different sheets
  - Add some border to make it easier to read
  - Simple styling for the header
- The vignette for `openxlsx` has **a lot** of examples of formatting!

## Write to Excel-file...
- Very simple formatting
```{r}
write.xlsx(list("EU groups" = group_untidy,
                "Countries" = mar_untidy),
           file = "output.xlsx",
           colNames = TRUE,
           borders = "columns",
           headerStyle = createStyle(fgFill = "yellow"))
```

## Exercise
- Create a new project
- Get the datasets from the shared drive and move them to your project-directory.
- Download the file `example.R`, put it in your project-directory and open it
- Run the code chunk by chunk and inspect the results.
- At line 92 there is a missing chunk that you should write yourselves. It is the bit that makes the group-data untidy so it can go into the excel-file
-  When you are done you can run the whole file by pressing "Run" in the top right
corner of the editor pane
