---
title: "Practical E - XXXX"
author: "Statistical Programming in R"
date: ""
output:
  html_document: 
    self_contained: false
    lib_dir: libs
  pdf_document: default
---


```{r echo=FALSE}
printAnswers <- TRUE
```

---

#### Exercises 

---

The following packages are required for this practical:

```{r, message=FALSE}
library(dplyr)
library(magrittr)
library(mice)
library(ggplot2)
```

and if you'd like the same results as I have obtained, you can fix the random seed

```{r cache=TRUE}
set.seed(123)
```

---

0. **Function `plot()` is the core plotting function in `R`. Find out more about `plot()`: Try both the help in the help-pane and `?plot` in the console. Look at the examples by running `example(plot)`.**

---

1. **Create a scatterplot between `age` and `bmi` in the `mice::boys` data set.**
Apply a standard plotting device in `R` or `ggplot`.
<!-- With a standard plotting device in `R`: -->

---

2. **Now recreate the plot with the following specifications:**

- If `bmi < 18.5` use `color = "light blue"`
- If `bmi > 18.5 & bmi < 25` use `color = "light green"`
- If `bmi > 25 & bmi < 30` use `color = "orange"`
- If `bmi > 30` use `color = "red"`


---

3. **Create a histogram for `age` in the `boys` data set.** 
Apply a standard plotting device in `R` or `ggplot`.

---

4. **Create a bar chart for `reg` in the boys data set**
Apply a standard plotting device in `R` or `ggplot`.
<!-- With a standard plotting device in `R`: -->

---

5. **Create a box plot for `hgt` with different boxes for `reg` in the `boys` data set**
Apply a standard plotting device in `R` or `ggplot`.
<!-- With a standard plotting device in `R`: -->

---

6. **Create a density plot for `age` with different curves for boys from the `city` and boys from rural areas (`!city`).** 
Apply a standard plotting device in `R` or `ggplot`.

---

7. **Create a diverging bar chart for `hgt` in the `boys` data set, that displays for every `age` year that year's mean height in deviations from the overall average `hgt`.** 
Apply `ggplot`.

---

8. **Read in the `sf` package, and open the shapefiles on the Danish municipalities from the course homepage. Plot the `REGIONNAVN` variable to see the Danish regions. Plot the municipal-level population.** Apply a standard plotting device in `R` or `ggplot`.


---

End of Practical

---

#### Useful References

[The `ggplot2` reference webpage](https://ggplot2.tidyverse.org/reference/index.html)


---


